Nonparametric Block Bootstrap Kolmogorov-Smirnov Goodness-of-Fit Test

### 2024-08-18 manuscript/manuscript.tex
minor comments on notation
use png figures for the qq plot with much smaller sizes
notation cleaning and proposition removal

Continuing using the notation in the independent case,
we use $F^{(b)}_n$ and $\hat\theta^{(b)}_n$ for the empirical distribution and
the estimated parameters based on the $b$th bootstrap sample,
$b = 1, \ldots, B$.
We then use block bootstrap to approximate the asymptotic distribution of
the KS statistic under $H_0$. Let $\hat{\theta}_0 = 
\mathbb{E}^{(b)}[\hat\theta^{(b)}_n]$. In particular, we
use the distribution of $\sqrt{n}(F^{(b)}_n(x) - \mathbb{E}^{(b)}[F^{(b)}_n(x)])$
as an approximation of the distribution of
$\sqrt{n}(F_n(x) - F(x; \theta_0))$, and the distribution of 
$\sqrt{n}(F(x; \hat\theta^{(b)}_n) - F(x; \hat{\theta}_0))$ as
an approximation of the distribution of 
$\sqrt{n}(F(x; \hat\theta_n) - F(x; \theta_0))$.
Here $\mathbb{E}^{(b)}$ denotes the expectation with respect to the bootstrap 
distribution 
(i.e., the randomness due
to the resampling using block bootstrap) conditional on the observations 
$X_1, \dots, X_n$.
The expected values $\mathbb{E}^{(b)}[F^{(b)}_n(x)]$ and
$\hat{\theta}_0 = \mathbb{E}^{(b)}[\hat\theta^{(b)}_n]$ can be approximated by, respectively,
$\mathbb{E}^{(b)}[F^{(b)}_n(x)] = \frac{1}{B}\sum_{b = 1}^BF^{(b)}_n(x)$ and
$\mathbb{E}^{(b)}[\hat\theta^{(b)}_n]  =  \frac{1}{B}\sum_{b = 1}^B\hat\theta^{(b)}_n$.

Then, we can define
\begin{align*}
  Y^{(b)}_n(x) &= \sqrt{n}(F^{(b)}_n(x) - \mathbb{E}^{(b)}[F^{(b)}_n(x)]) - 
             \sqrt{n}(F(x; \hat\theta^{(b)}_n) - F(x; \hat{\theta}_0) \\
           &= \sqrt{n}(F^{(b)}_n(x) - F(x; \hat\theta^{(b)}_n)) -
             \sqrt{n}(\mathbb{E}^{(b)}[F^{(b)}_n(x)] - F(x; \hat{\theta}_0)),
\end{align*}
and $T^{(b)}_n = \sup_x|Y^{(b)}_n(x)|$. Each $T_n^{(b)}$,
$b =1, \ldots, B$, is considered a draw from a distribution that approximates
the distribution of $T_n$. Therefore, the p-value of the observed statistic
$T_n$ can be assessed by positioning it against the empirical distribution of
$T_n^{(b)}$, $b = 1, \ldots, B$.

In summary, the procedure of the NPBB test is 
summarized as follows. Repeat the following steps for $b \in \{1, ..., B\}$.
\begin{enumerate}
\item
  Generate $X^{(b)}_1,...,X^{(b)}_n$ by applying moving block bootstrap 
  on the original sample as
  defined previously.
\item
  Obtain parametrically fitted parameters 
  $\hat\theta^{(b)}_n$ of $\theta$ from $X^{(b)}_1,...,X^{(b)}_n$.  
\item
  Obtain the empirical distribution function $F^{(b)}_n$ of
  $X^{(b)}_1,...,X^{(b)}_n$. 
\item
  Calculate bootstrap KS statistic
  \[
    T^{(b)}_n = \sup_x \Big\vert \sqrt{n}\left(F^{(b)}_n(x) 
    - F(x; \hat\theta^{(b)}_n)\right) - B_n(x) \Big\vert.
  \]
  where 
  $B_{n}(x) = \sqrt{n}(\mathbb{E}^{(b)}[F^{(b)}_n(x)] - 
  F(x; \hat{\theta}_0))$ is the known
  bias term.
\end{enumerate}
The p-value of the block bootstrap KS test can be approximated
as $p = \sum_{b=1}^B I\{T^{(b)}_n > T_n\} / B$.

Assume that the time series $\{X_i\}$ is stationary and
strongly mixing. The following results provide theoretical validity
of the NPBB procedure of the KS test. Sketches of their proofs are
in the Supporting Information, with most arguments similar to those
in \citet{kunsch1989jackknife}.

\begin{proposition}\label{prop:Tn}
  As $1/l+l/n\rightarrow 0$,
  \[
    T_n \overset{d}{\rightarrow}
    \sup_x \big\vert G_{\infty}(x)+f(x;\theta_0)U \big\vert
  \]
  for some appropriately defined Gaussian process $G_{\infty}(\cdot)$
  and Gaussian random variable  $U$.
\end{proposition}

\begin{proposition}\label{prop:Tnboot}
  As $1/l+l/n\rightarrow 0$,
  \[
    \big\vert \mathbb{P}(T_n^{(b)}\leq x \mid \{X_i: i = 1, \ldots, n\}) -
    \mathbb{P}(T_n\leq x) \big\vert
    \overset{a.s.}{\rightarrow} 0
  \]
  for any fixed $x$, where $T_n^{(b)}$ is the statistic computed 
  based on the bootstrap sample $\{X_i^{(b)}: i = 1, \ldots, n\}$.
\end{proposition}

Following the arguments in the proof of the Glivenko--Cantelli
Theorem, the result in Proposition~\ref{prop:Tnboot} can be further
strengthened to allow uniform convergence over $x$. That is,
as $1/l+l/n\rightarrow 0$,
\[
  \sup_x \big \vert
  \mathbb{P}(T_n^{(b)}\leq x \mid \{X_i: i = 1, \ldots, n\}) - \mathbb{P}(T_n\leq x) \big\vert
  \overset{a.s.}{\rightarrow} 0.
\]

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width = .9\textwidth]{figures/normal}
  \includegraphics[width = .9\textwidth]{figures/gamma}
  \caption{Q-Q plots of the p-values testing that a time series
    have marginal normal distribution with true data generating distribution
    being $N(8,8)$ (upper) and marginal gamma distribution with true data
    generating distribution $\Gamma(8,1)$ (lower).}
  \label{fig:qq}
\end{figure}

@@

The KS test remains a crucial tool for researchers across diverse fields,
enabling critical assessments of population distributions. We introduced a
NPBB variant of the KS test tailored for data with
serial dependencies, which eliminates the need for explicit modeling serial
dependence in addition to parameter specification within the hypothesized
distribution. Our simulations have demonstrated that, given
sufficiently large samples of time series data and moderate serial dependencies,
the test maintains its size under the null hypothesis and have considerable
power in detecting deviations from the hypothesized marginal distribution. The
method's performance improves as the strength of temporal dependence decreases
or the sample size increases. For moderate serial dependence levels in practice,
the test can be recommended for sample size~$n = 400$ or higher.
Overall, the test provides an analog of the bias-corrected NPB KS test of 
\citet{babu2004goodness} in the context of stationaly time
series with unspecified serial dependence. Compared to the SPB method
of \citet{zeimbekakis2022misuses}, which requires a working dependence model, it
is reliable regardless of whether the working dependence model approximates the
true dependence model.

Several promising directions for further research emerge from the findings of
this study. Firstly, analogous methodologies for other goodness-of-fit tests
based on empirical distributions, such as the Anderson--Darling and Cramér--von
Mises tests---which have been shown to outperform the KS test in certain scenarios
\citep{stephens2017tests}---could be explored. The application of block
bootstrapping bias correction to these tests could be particularly beneficial in
environmental studies, such as extreme value modeling, where observations are
often serially dependent. Secondly, an extension of the proposed NPBB
one-sample KS test to a two-sample setting would be
intriguing. This adaptation would test whether two samples originate from the
same distribution without requiring specification of serial dependencies in
either sample, maintaining the advantageous feature of the original method.

@@

% \nocite{*}% Show all bib entries - both cited and uncited; comment this line to view only cited bib entries;

\section*{Supporting Information}

The proofs of the two propositions in \ref{sub:npbb} are sketched in
the supplement.

>>>>

As in the independent case,
we use $F^{(b)}_n$ and $\hat\theta^{(b)}_n$ for the empirical distribution and
the estimated parameters based on the $b$th bootstrap sample,
$b = 1, \ldots, B$.
We then use block bootstrap to approximate the asymptotic distribution of
the KS statistic under $H_0$. Let
$\theta_0^* = \mathbb{E}^{*}[\hat\theta^{(b)}_n]$, where
$\mathbb{E}^{*}$ denotes the expectation with respect to the
bootstrap distribution (i.e., the randomness due
to the resampling using block bootstrap) conditional on the observations 
$X_1, \dots, X_n$. In particular, we
use the distribution of $\sqrt{n}(F^{(b)}_n(x) - \mathbb{E}^{*}[F^{(b)}_n(x)])$
as an approximation of the distribution of
$\sqrt{n}(F_n(x) - F(x; \theta_0))$, and the distribution of 
$\sqrt{n}(F(x; \hat\theta^{(b)}_n) - F(x; \theta_0^*))$ as
an approximation of the distribution of 
$\sqrt{n}(F(x; \hat\theta_n) - F(x; \theta_0))$.
Here, the expected values $\mathbb{E}^{*}[F^{(b)}_n(x)]$ and
$\theta_0^* = \mathbb{E}^{*}[\hat\theta^{(b)}_n]$ can be
approximated by, respectively,
$\mathbb{E}_B^{*}[F^{(b)}_n(x)] = \frac{1}{B}\sum_{b = 1}^B F^{(b)}_n(x)$ and
$\mathbb{E}_B{*}[\hat\theta^{(b)}_n]  =  \frac{1}{B}\sum_{b = 1}^B\hat\theta^{(b)}_n$.

Then, for each $b \in \{1, \dots, B\}$, we can define
\begin{align*}
  Y^{(b)}_n(x) &= \sqrt{n}(F^{(b)}_n(x) - \mathbb{E}^{*}[F^{(b)}_n(x)]) - 
             \sqrt{n}(F(x; \hat\theta^{(b)}_n) - F(x; \theta_0^*) \\
           &= \sqrt{n}(F^{(b)}_n(x) - F(x; \hat\theta^{(b)}_n)) -
             \sqrt{n}(\mathbb{E}^{*}[F^{(b)}_n(x)] - F(x; \theta_0^*)),
\end{align*}
and $T^{(b)}_n = \sup_x|Y^{(b)}_n(x)|$. Each $T_n^{(b)}$,
$b =1, \ldots, B$, is considered a draw from a distribution that approximates
the distribution of $T_n$. Therefore, the p-value of the observed statistic
$T_n$ can be assessed by positioning it against the empirical distribution of
$T_n^{(b)}$, $b = 1, \ldots, B$.

In summary, the procedure of the NPBB test is 
summarized as follows. Repeat the following steps for $b \in \{1, ..., B\}$.
\begin{enumerate}
\item
  Generate $X^{(b)}_1,...,X^{(b)}_n$ by applying moving block bootstrap 
  on the original sample as
  defined previously.
\item
  Obtain parametrically fitted parameters 
  $\hat\theta^{(b)}_n$ of $\theta$ from $X^{(b)}_1,...,X^{(b)}_n$.  
\item
  Obtain the empirical distribution function $F^{(b)}_n$ of
  $X^{(b)}_1,...,X^{(b)}_n$. 
\item
  Calculate bootstrap KS statistic
  \[
    T^{(b)}_n = \sup_x \Big\vert \sqrt{n}\left(F^{(b)}_n(x) 
    - F(x; \hat\theta^{(b)}_n)\right) - B_n(x) \Big\vert.
  \]
  where 
  $B_{n}(x) = \sqrt{n}(\mathbb{E}^{*}[F^{(b)}_n(x)] - 
  F(x; \theta_0^*))$ is the approximated bias term.
\end{enumerate}
The p-value of the block bootstrap KS test can be approximated
as $p = \sum_{b=1}^B I\{T^{(b)}_n > T_n\} / B$.

Assume that the time series $\{X_i\}$ is stationary and
strongly mixing. Using arguments similar to those
in \citet{kunsch1989jackknife}, we conjecture the following results
justifying the validity of the NPBB procedure of the KS test.
As $1/l+l/n\rightarrow 0$,
\[
  T_n \overset{d}{\rightarrow}
  \sup_x \big\vert G_{\infty}(x)+f(x;\theta_0)U \big\vert
\]
for some appropriately defined Gaussian process $G_{\infty}(\cdot)$
and Gaussian random variable  $U$.
Further,  as $1/l+l/n\rightarrow 0$,
\[
  \big\vert \mathbb{P}(T_n^{(b)}\leq x \mid \{X_i: i = 1, \ldots, n\}) -
  \mathbb{P}(T_n\leq x) \big\vert
  \overset{a.s.}{\rightarrow} 0
\]
for any fixed $x$, where $T_n^{(b)}$ is the statistic computed 
based on the bootstrap sample $\{X_i^{(b)}: i = 1, \ldots, n\}$.
Following the arguments in the proof of the Glivenko--Cantelli
Theorem, the bootstrap consistency result could be further
strengthened to allow uniform convergence over $x$. That is,
as $1/l+l/n\rightarrow 0$,
\[
  \sup_x \big \vert
  \mathbb{P}(T_n^{(b)}\leq x \mid \{X_i: i = 1, \ldots, n\}) - \mathbb{P}(T_n\leq x) \big\vert
  \overset{a.s.}{\rightarrow} 0.
\]

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width = .9\textwidth]{figures/normal.png}
  \includegraphics[width = .9\textwidth]{figures/gamma.png}
  \caption{Q-Q plots of the p-values testing that a time series
    have marginal normal distribution with true data generating distribution
    being $N(8,8)$ (upper) and marginal gamma distribution with true data
    generating distribution $\Gamma(8,1)$ (lower).}
  \label{fig:qq}
\end{figure}

@@

The KS test remains a crucial tool for researchers across diverse fields,
enabling critical assessments of population distributions. We introduced a
NPBB variant of the KS test tailored for data with
serial dependencies, which eliminates the need for explicit modeling serial
dependence in addition to parameter specification within the hypothesized
distribution. Our simulations have demonstrated that, given
sufficiently large samples of time series data and moderate serial dependencies,
the test maintains its size under the null hypothesis and have considerable
power in detecting deviations from the hypothesized marginal distribution. The
method's performance improves as the strength of temporal dependence decreases
or the sample size increases. For moderate serial dependence levels in practice,
the test can be recommended for sample size~$n = 400$ or higher.
Overall, the test provides an analog of the bias-corrected NPB KS test of 
\citet{babu2004goodness} in the context of stationaly time
series with unspecified serial dependence. Compared to the SPB method
of \citet{zeimbekakis2022misuses}, which requires a working dependence model, it
is reliable with no need to worry about whether the working dependence
model approximates the true dependence model.

Several promising directions for further research emerge from the findings of
this study. Firstly, analogous methodologies for other goodness-of-fit tests
based on empirical distributions, such as the Anderson--Darling and Cramér--von
Mises tests---which have been shown to outperform the KS test in certain scenarios
\citep{stephens2017tests}---could be explored. The application of block
bootstrapping bias correction to these tests could be particularly beneficial in
environmental studies, such as extreme value modeling, where observations are
often serially dependent. Secondly, an extension of the proposed NPBB
one-sample KS test to a two-sample setting would be
intriguing. This adaptation would test whether two samples originate from the
same distribution without requiring specification of serial dependencies in
either sample, maintaining the advantageous feature of the original method.
Finally, a rigorous theoretical justification of the NPBB method for
stationary series is of significant academic interest. Establishing
precise conditions for the asymptotic behavior of the KS statistic and
the consistency of the bootstrap procedure would offer valuable
practical insights into the scenarios where the NPBB approach is
expected to perform optimally.

@@
